---
title: "Introduction to __DSArray__"
author: "Peter Hickey"
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
#abstract: >
#  Vignette Abstract
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

__TODO:__ Use simpler, small data to demonstrate functions because these allow you to actually see results of the operations. Save example data for performance 
vignette. 

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(DSArray)
```

__DSArray__ ("_desiree_") provides efficient in-memory representation of 3-dimensional arrays that contain many duplicate slices via the _DSArray_ (**D**uplicate **S**lice **Array**) S4 class. A basic array-like API is provided for instantiating, subsetting, and combining DSArray objects.

This vignette introduces the _DSArray_ class and demonstrates common operations. The [benchmarking](benchmarking.html) vignette compares the use of _DSArray_ objects, _array_ objects from the [__base__](https://stat.ethz.ch/R-manual/R-patched/library/base/html/00Index.html) package, sparse matrix objects from the `r CRANpkg("Matrix")` package, and _HDF5Array_ objects from the `r Biocpkg("HDF5Array")` package.
 
## What the hell do I do with this?

The __DSArray__ package serves a niche purpose. However, since I've found it useful, I'm making it publicly available. Here is the problem in words and a picture illustrating the solution that `r Biocpkg("DSArray")` offers.

Suppose you have data on a set of `n` samples where each sample's data can be represented as a matrix (`x1`, ..., `xn`) where `dim(x1) = ... = dim(xn) = c(nrow, ncol)`. We can combine these matrices along a given dimension to form a 3-dimensional array, `x`. __DSArray__ is designed for the special case where there are many duplicate _slices_ of `x`. Continuing our example, if each of the `x1`, ..., `xn` have duplicate rows and we combine `x1`, ..., `xn` to form `x` such that `x[, j, ]` represents `xj`, then for this special case we can efficiently represent the data by storing only the unique rows of the `x1`, ..., `xn` and an associated index. A picture will hopefully help make this clearer:

```{r echo = FALSE, draw-DSArray}
set.seed(666)
# TODO: Figure doesn't look the same in vignette as it does in RStudio
n <- 3
nrow <- 20
ncol = 8
DSArray:::.drawDSArray(n = 3, nrow = 20, ncol = 8)
```

In this example we have `n = 3` matrices, each shown as a slice of `x` (`x[, 1, ]`, `x[, 2, ]`, `x[, 3, ]`) with `nrow = 20` and `ncol = 8`, where the colour of the row identifies identical rows. Note that the same row may be found multiple times within a sample and may also be common to multiple samples. We can construct the _DSArray_ representation of `x` by calling `DSArray(x)`. The _DSArray_ representation has a _key_ and a _val_, much like an associative array, map, or dictionary. The j-th column of the _key_ is the key for the j-th sample (note the colour ordering of each sample). The _val_ contains all __unique__ rows found in the `n` samples.

We can reconstruct the data for a particular sample by expanding the _val_ by the relevant column of the _key_. We can often compute the required summaries of the data while retaining this sparse representation. In this way, a _DSArray_ is similar to using a run length encoding of a vector or a sparse matrix representation to leverage the additional structure in the object.

## Constructing a _DSArray_

The `DSArray()` function provides several different ways to construct a _DSArray_ object. We demonstrate it's use when working with a single sample or multiple samples. 

### Single sample

Here we have data from a single sample in a _matrix_, from which we wish to construct the _DSArray_ representation:

```{r, matrix-input}
m <- matrix(1:10, ncol = 2, dimnames = list(letters[1:5], LETTERS[1:2]))
m 
m_dsa <- DSArray(m)
m_dsa
```

Note that columns of `m` becomes the _slices_ of `m_dsa`; this is because a _DSArray_ uses columns to represent samples. Also note that by default `DSArray()` constructs the `dimnames` from the input:

```{r, matrix-input-dimnames}
dimnames(m)
dimnames(m_dsa)
```

We can override these by supplying them as the `dimnames` argument, in particular to set the column names (sample names) for these data:

```{r, matrix-input-set-dimnames}
dimnames(DSArray(m, dimnames = list(rownames(m), "sample-1", colnames(m))))
```

### Multiple samples

When we have data on multiple samples, these might already be represented as a 3-dimensional _array_:

```{r, array}
a <- array(c(1, 3, 5, 10, 30, 50, 100, 300, 500, 2, 4, 6, 20, 40, 60, 200, 
             400, 600),
           dim = c(3, 3, 2),
           dimnames = list(letters[1:3], LETTERS[1:3], letters[25:26]))
```

Here, each sample's data are a column of `a`:

```{r, array-input-single-sample}
# Sample A
a[, "A", ]
# Sample B
a[, "B", ]
# Sample C
a[, "C", ]
```

By default, the `DSArray()` constructor assumes the columns of an _array_ input represent the samples:

```{r, array-input}
a_dsa <- DSArray(a)
a_dsa
```

But we can specify this explicitly by setting the `MARGIN` argument:

```{r, array-input-margin}
# Default: Columns (2) as samples
DSArray(a, MARGIN = 2)
# Rows (1) as samples
DSArray(a, MARGIN = 1)
# Slices (3) as samples
DSArray(a, MARGIN = 3)
```

Alternatively, the data may be represented as a _list_ of _matrix_ objects, one per sample, where the dimensions of each _matrix_ are identical:

```{r, list-input}
l <- list(A = a[, "A", ],
          B = a[, "B", ],
          C = a[, "C", ])
l
l_dsa <- DSArray(l)
l_dsa
```

### Example data

The above toy examples demonstrate how to construct a _DSArray_, but these data are too simple to benefit from a _DSArray_ representation. For the remainder of this vignette we will use the `dsa` example dataset included with __DSArray__:

```{r, load-data}
dsa
```

While these data are extracted from real biological experiments, it is not necessary to understand the details to follow the vignette. Briefly, `dsa` contains counts of patterns of DNA methylation at 1,000,000 randomly sampled features (rows) from 17 samples (columns). For each feature and sample, the number of occurences of each of the $2^4 = 16$ observable patterns of DNA methylation is recorded (slices).

Importantly, these data are very sparse. Although at least one sample has at least one observation for every feature, the majority of samples have no observations for the majority of features (these are recorded as `NA`). Furthermore, there are many features and samples with identical ("duplicated") counts, hence these data are perfectly suited to storing as a _DSArray_:

```{r, dsa-summary}
# TODO: Write a summary,DSArray-method; might be interesting to know percentage 
#       NA for each sample and summary() of non-NA values per sample/slice
```

We can compute the memory saving of using a _DSArray_ representation rather than an _array_[^pryr]. We find that the _DSArray_ representation is `r as.numeric(round(100 * pryr::object_size(dsa) / pryr::object_size(as(dsa, "array"))), 0)`% the size of the equivalent _array_ representation:

[^pryr]: This uses the `object_size()` function from the `r CRANpkg("pryr")` package.

```{r, memory-saving}
pryr::object_size(dsa)
# Create the array representation
a <- as(dsa, "array")
pryr::object_size(a)
```

In what follows, we will demonstrate that working with the _DSArray_ respresenatation, `dsa`, gives equivalent results to using the _array_ representation, `a`, with the benefit of much reduced memory usage.

## API and overview of methods

The aim is to allow a _DSArray_ to be used as a drop-in replacement for an _array_ when the need arises. The _DSArray_ API is therefore written to mimic the _array_ API so that _DSArray_ objects behave as if they were 3-dimensional _array_ objects. However, the API coverage is not 100% complete. I am adding these missing methods as needed, so if something you require is missing then please get in touch by filing a feature request at [https://github.com/PeteHaitch/DSArray/issues](https://github.com/PeteHaitch/DSArray/issues). 

Where possible, __DSArray__ aims to avoid "densifying" the data (i.e. converting it to an _array_ for intermediate calculations) since doing so obviously negates the memory efficiency of using a _DSArray_. However, not all operations are optimally implemented, some because they are difficult (or perhaps impossible) and others because I haven't yet taken the time to optimise them. 

As an example, one operation to avoid if at all possible is subset replacement with the `[<-` operator; this is a very expensive operation since it first densifies the data and then re-sparsifies. 

Thus a _DSArray_ may best be thought of as a read-only data structure, at least in its current form.

### Subsetting 

We can subset a _DSArray_ just as we would an _array_ by using the `[` operator[^drop]:

[^drop]: The `drop` argument to `[` is always set to `FALSE` when subsetting a _DSArray_.

```{r, [}
# Extract the first feature
dsa[1, , ]
# Extract the first sample
dsa[, 1, ]
# Extract the first slice
dsa[, , 1]
# Extract the first 10 features for the first 3 samples
# TODO: This is broken
# dsa[1:10, 1:3, ]
```

<!--TODO: Comment on the speed of subsetting by different combinations of i, j, and k; link to benchmarking vignette -->

### Combining

__TODO__

- `acbind()`
- `arbind()`

### Arithmetic, Comparison, and Logic

__TODO:__ Perhaps use simpler, small data to demonstrate functions because these allow you to actually see results of the operations. Also, implement a show method that includes head/tail of data ala HDF5Array, which will help in this regard.

It is trivial to implement high-performance scalar-_DSArray_ arithmetic, comparison, and logic methods:

```{r, scalar-ops}
# Arithmetic
dsa + 3
dsa - 3
dsa * 3
dsa ^ 3
dsa %% 3
dsa %/% 3
dsa / 3

# Comparison
dsa == 3
dsa > 3
dsa < 3
dsa != 3
dsa <= 3
dsa >= 3

# Logic
dsa & TRUE
dsa | TRUE
```

The same is not true of vector-_DSArray_, array-_DSArray_, or _DSArray_-_DSArray_ operations, which all currently require the densification of the _DSArray_ argument(s):

```{r, array-ops}
dsa + 1:2
```

We can check that the _DSArray_ method gives an identical result to the _array_ method using the non-exported `DSArray:::dsa_identical_to_array()` function. For example:

```{r, dsa_identical_to_array}
DSArray:::dsa_identical_to_array(dsa * 3, a * 3)
```

### Math and Math2

__UP TO HERE__

### Summary

### Complex

## Using _DSArray_ within a _SummarizedExperiment_

An efficient representation of sparse 3-dimensional arrays within a _SummarizedExperiment_ was the motivation for the __DSArray__ project. The __SummarizedExperiment__ package defines an important base class in the Bioconductor project. I needed an efficient way to store DNA methylation patterns, a particular kind of genomic data, that was compatible with the __SummarizedExperiment__ package.

## Future work
